---
title: "Longitudinal PACC Analyses"
date: "`r Sys.Date()`"
output: 
  html_document:
     prefer-html: true
---

```{r setup, include = FALSE, warning = FALSE, message = FALSE}
library(plyr)
library(nlme)
library(ggplot2)
library(sjPlot)
library(dplyr)
library(gtsummary)
library(flextable)
library(ggrepel)
library(readr)
library(DT)
library(tidyr)

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
theme_set(theme_bw())
```

```{r}
load(file = "~/Dropbox (Partners HealthCare)/Mabel/Longitudinal A4 & Gene Expression/A4_GeneExp_Project/Data/RNAseqData_Autosomal_XLinked_wDose_120424.RData")

gene_names <- readRDS("~/Dropbox (Partners HealthCare)/Mabel/Longitudinal A4 & Gene Expression/A4_GeneExp_Project/Data/A4_RNAseq_GeneNames_081424.rds")

## PACC Models
load("~/Dropbox (Partners HealthCare)/Mabel/Longitudinal A4 & Gene Expression/A4_GeneExp_Project/Data/PACC Models/LME_PACC_GeneSexE4_Models_2024-12-06.RData")
load("~/Dropbox (Partners HealthCare)/Mabel/Longitudinal A4 & Gene Expression/A4_GeneExp_Project/Data/PACC Models/LME_PACC_GeneAb_Models_2024-12-05.RData")
load("~/Dropbox (Partners HealthCare)/Mabel/Longitudinal A4 & Gene Expression/A4_GeneExp_Project/Data/PACC Models/LME_PACC_GeneSexAb_Models_2024-12-06.RData")

# ## Amyloid Models
# load("~/Dropbox (Partners HealthCare)/Mabel/Longitudinal A4 & Gene Expression/Amyloid Models/LME_Ab_GeneE4_Models_2024-12-07.RData")


```

```{r}
gene_names <- select(gene_names, gene = ensembl_gene_id, chromosome_name, external_gene_name)
```

```{r}

### Gather all baseline data for demo table (baseline is not the same across all measures)

baseline <- rna %>%
  group_by(BID) %>%
  arrange(BID, pacc_time) %>%
  slice(1)

bl_amyloid <- rna %>%
  filter(!is.na(suvr_cer)) %>%
  group_by(BID) %>%
  arrange(BID, pacc_time) %>%
  slice(1) %>%
  select(BID, bl_Amyloid = suvr_cer, bl_AmyloidStatus = overall_score)

baseline <- merge(baseline, bl_amyloid, by = "BID", all.x = TRUE)

## Calculate follow up 

pacc_followup <- rna %>%
  group_by(BID) %>%
  arrange(BID, pacc_time) %>%
  slice_tail() %>%
  select(BID, pacc_followup = pacc_time)

baseline <- merge(baseline, pacc_followup, by = "BID", all.x = TRUE)

baseline <- baseline %>%
  mutate(pacc_followup = case_when(
    pacc_followup == 0 ~ NA,
    TRUE ~ pacc_followup
  ))

```

# Significant Results Table

Summary table of all significant results (after fdr correction) from the following longitudinal PACC analyses:

1.  PACC \~ GeneExp
2.  PACC \~ GeneExp\*Sex
3.  PACC \~ GeneExp\*E4
4.  PACC \~ GeneExp\*Sex\*E4
5.  PACC \~ GeneExp\*Ab
6.  PACC \~ GeneExp\*Ab\*Sex

```{r}
PACCTable_GeneSexE4_wDose$predictor <- "Gene*Sex*E4"
PACCTable_GeneAb_wDose$predictor <- "Gene*Ab"
PACCTable_GeneSexAb_wDose$predictor <- "Gene*Sex*Ab"

All_Genes <- rbind.fill(PACCTable_GeneSexE4_wDose, 
                        PACCTable_GeneAb_wDose, PACCTable_GeneSexAb_wDose)

All_Genes_Sig <- filter(All_Genes, fdr_p < 0.05)
all_sig_genes <- unique(All_Genes_Sig$gene)
```

```{r}
datatable(All_Genes_Sig,
          filter = "top",
          rownames = FALSE,
          extensions = c('Buttons'),
          options = list(dom = 'Btip',
                         buttons = c('copy', 'csv'),
                         pageLength = 15,
                         columnDefs = list(list(className = 'dt-center', targets = "_all")))) %>%
  formatRound(columns = c("beta","se", "t", "p", "fdr_p"), digits = 4) %>% 
  formatStyle("fdr_p", color = styleInterval(0.05, c("red", "black")))
```

```{r}

```

# Analyses {.tabset}

All models are adjusted for sex, baseline age, years of education, cohort/treatment (A4 Placebo, A4 Treated, LEARN/SF), PACC story version, and cumulative dose. PACC story version and cumulative dose are time varying variables.

## PACC \~ Gene x Sex x APOEe4 {.tabset}

```{r}
PACCTable_GeneSexE4 <- PACCTable_GeneSexE4_wDose %>%
  left_join(gene_names, by = "gene") %>%
  mutate(Sig = case_when(
    p < 0.05 ~ T,
    TRUE ~ F
  ))
```

There is `r length(subset(PACCTable_GeneSexE4, fdr_p < 0.05)$gene)` significant gene that survives FDR correction.

### Volcano Plot

```{r}
sig_bon <- -log10(0.05/length(unique(PACCTable_GeneSexE4$gene))) #Bonferroni correction for plotting

ggplot(PACCTable_GeneSexE4, aes(x = beta, y = -log10(p), color = Sig)) +
  geom_point() +
  geom_hline(yintercept = -log10(0.05), color = "red", linetype = "dashed", size = 1.2) +
  geom_hline(yintercept = sig_bon, color = "blue", linetype = "dashed", size = 1) +
  theme(legend.position = "none") +
  scale_colour_manual(values = setNames(c('seagreen','black'), c(T,F))) +
  geom_text(x = -6, y = 0.75, label = "p = 0.05", check_overlap = TRUE, color = "black") +
  geom_text(x = -6, y = (sig_bon-0.5), label = "p = 0.05 \n Bonferroni corrected", check_overlap = TRUE, color = "black") +
  geom_label_repel(aes(label = ifelse(-log10(p) > 6 | beta < -5, gene, "")), 
                   nudge_x = -0.05, nudge_y = 0.04, min.segment.length = 0) +
  labs(
    title = "PACC ~ Gene*Sex*APOEe4 Non-Adjusted P-Values"
  )

# plot without outlier
ggplot(subset(PACCTable_GeneSexE4, beta > -5), aes(x = beta, y = -log10(p), color = Sig)) +
  geom_point() +
  geom_hline(yintercept = -log10(0.05), color = "red", linetype = "dashed", size = 1.2) +
  geom_hline(yintercept = sig_bon, color = "blue", linetype = "dashed", size = 1) +
  theme(legend.position = "none") +
  scale_colour_manual(values = setNames(c('seagreen','black'), c(T,F))) +
  geom_text(x = (-0.85*max(PACCTable_GeneSexE4$beta)), y = (-log10(0.05)-0.25), label = "p = 0.05", check_overlap = TRUE, color = "black") +
  geom_text(x = (-0.75*max(PACCTable_GeneSexE4$beta)), y = (sig_bon-0.5), label = "p = 0.05 \n Bonferroni corrected", check_overlap = TRUE, color = "black") +
  geom_label_repel(aes(label = ifelse(-log10(p) > 6 | beta < -5, gene, "")), 
                   nudge_x = -0.05, nudge_y = 0.04, min.segment.length = 0) +
  labs(
    title = "PACC ~ Gene*Sex*APOEe4 Non-Adjusted P-Values"
  )
```

### Summary Table

```{r}
datatable(subset(PACCTable_GeneSexE4, Sig == T),
          filter = "top",
          rownames = FALSE,
          extensions = c('Buttons'),
          options = list(dom = 'Btip',
                         buttons = c('copy', 'csv'),
                         pageLength = 15,
                         columnDefs = list(list(className = 'dt-center', targets = "_all")))) %>%
  formatRound(columns = c("beta","se", "t", "p", "fdr_p"), digits = 4) %>%
  formatStyle("fdr_p", color = styleInterval(0.05, c("red", "black")))
```

```{r}
rna$SEX <- as.factor(rna$SEX)
rna$APOE <- as.factor(rna$APOE)
rna$Cohort <- as.factor(rna$Cohort)
rna$QSVERSION <- as.factor(rna$QSVERSION)
```

### Significant Results Plots {.tabset}

::: panel-tabset
#### ENSG00000211745

```{r, fig.width = 12}
fit <- lme(PACC ~ ENSG00000211745*SEX*APOE*pacc_time + bl_pacc_age*pacc_time +
             EDCCNTU*pacc_time + Cohort*pacc_time + QSVERSION + Cumulative_Dose_Scaled, 
           data = rna, random =~ pacc_time | BID, na.action = "na.omit", 
           control = list(opt = 'optim', msMaxIter = 1000, msMaxEval = 1000))


plot_model(fit, type = "pred", terms = c("pacc_time", "ENSG00000211745", "SEX", "APOE")) +
  #theme(strip.background = element_rect(fill = "white")) +
  labs(
    x = "Time (Years)",
    y = "Longitudinal PACC",
    title = bquote("Longitudinal PACC by TRBV4-2, Sex, and APOEε4"),
    color = "ENSG00000233785\n(TRBV4-2)"
  )
```
:::

## PACC \~ Gene x Amyloid {.tabset}

```{r}
PACCTable_GeneAb <- PACCTable_GeneAb_wDose %>%
  left_join(gene_names, by = "gene") %>%
  mutate(Sig = case_when(
    p < 0.05 ~ T,
    TRUE ~ F
  ))
```

There are `r length(subset(PACCTable_GeneAb, fdr_p < 0.05)$gene)` significant gene that survives FDR correction.

### Volcano Plot

```{r}
sig_bon <- -log10(0.05/length(unique(PACCTable_GeneAb$gene))) #Bonferroni correction for plotting

ggplot(PACCTable_GeneAb, aes(x = beta, y = -log10(p), color = Sig)) +
  geom_point() +
  geom_hline(yintercept = -log10(0.05), color = "red", linetype = "dashed", size = 1.2) +
  geom_hline(yintercept = sig_bon, color = "blue", linetype = "dashed", size = 1) +
  theme(legend.position = "none") +
  scale_colour_manual(values = setNames(c('seagreen','black'), c(T,F))) +
  geom_text(x = (0.92*max(PACCTable_GeneAb$beta)), y = (-log10(0.05)-0.25), label = "p = 0.05", check_overlap = TRUE) +
  geom_text(x = (0.7*max(PACCTable_GeneAb$beta)), y = (sig_bon-0.5), label = "p = 0.05 \n Bonferroni corrected", check_overlap = TRUE) +
  geom_label_repel(aes(label = ifelse(-log10(p) > 5.5, gene, "")), 
                   nudge_x = -0.5, nudge_y = 0.05, min.segment.length = 0) +
  labs(
    title = "PACC ~ Gene*Amyloid Burden Non-Adjusted P-Values"
  )
```

### Summary Table

```{r}
datatable(subset(PACCTable_GeneAb, Sig == T),
          filter = "top",
          rownames = FALSE,
          extensions = c('Buttons'),
          options = list(dom = 'Btip',
                         buttons = c('copy', 'csv'),
                         pageLength = 15,
                         columnDefs = list(list(className = 'dt-center', targets = "_all")))) %>%
  formatRound(columns = c("beta","se", "t", "p", "fdr_p"), digits = 4) %>%
  formatStyle("fdr_p", color = styleInterval(0.05, c("red", "black")))
```

```{r}
bl_amyloid <- rna %>%
  filter(!is.na(suvr_cer)) %>%
  group_by(BID) %>%
  arrange(BID, pacc_time) %>%
  slice(1) %>%
  select(BID, bl_Amyloid = suvr_cer, bl_AmyloidStatus = overall_score)

rna <- merge(rna, bl_amyloid, by = "BID")
```

### Significant Results Plots {.tabset}

Plots of select significant genes

::: panel-tabset
#### ENSG00000115942

```{r, fig.width = 12}
fit <- lme(PACC ~ ENSG00000115942*bl_Amyloid*pacc_time + SEX*pacc_time + bl_pacc_age*pacc_time + EDCCNTU*pacc_time + Cohort*pacc_time + QSVERSION + Cumulative_Dose_Scaled, 
           data = rna, random =~ pacc_time | BID, na.action = "na.omit", 
           control = list(opt = 'optim', msMaxIter = 1000, msMaxEval = 1000))


plot_model(fit, type = "pred", terms = c("pacc_time", "ENSG00000115942", "bl_Amyloid")) +
  theme(strip.background = element_rect(fill = "white")) +
  labs(
    x = "Time (Years)",
    y = "Longitudinal PACC",
    title = bquote("Longitudinal PACC by ORC2 and Amyloid Status"),
    color = "ENSG00000115942\n(ORC2)"
  )
```

#### ENSG00000240132

```{r, fig.width = 12}
fit <- lme(PACC ~ ENSG00000240132*bl_Amyloid*pacc_time + SEX*pacc_time + bl_pacc_age*pacc_time + EDCCNTU*pacc_time + Cohort*pacc_time + QSVERSION + Cumulative_Dose_Scaled, 
           data = rna, random =~ pacc_time | BID, na.action = "na.omit", 
           control = list(opt = 'optim', msMaxIter = 1000, msMaxEval = 1000))


plot_model(fit, type = "pred", terms = c("pacc_time", "ENSG00000240132", "bl_Amyloid")) +
  theme(strip.background = element_rect(fill = "white")) +
  labs(
    x = "Time (Years)",
    y = "Longitudinal PACC",
    title = bquote("Longitudinal PACC by ETF1P2 and Amyloid Status"),
    color = "ENSG00000240132\n(ETF1P2)"
  )
```
:::

## PACC \~ Gene x Sex x Amyloid {.tabset}

```{r}
PACCTable_GeneSexAb <- PACCTable_GeneSexAb_wDose %>%
  left_join(gene_names, by = "gene") %>%
  mutate(Sig = case_when(
    p < 0.05 ~ T,
    TRUE ~ F
  ))
```

There are `r length(subset(PACCTable_GeneSexAb, fdr_p < 0.05)$gene)` significant gene that survives FDR correction.

### Volcano Plot

```{r}
sig_bon <- -log10(0.05/length(unique(PACCTable_GeneSexAb$gene))) #Bonferroni correction for plotting

ggplot(PACCTable_GeneSexAb, aes(x = beta, y = -log10(p), color = Sig)) +
  geom_point() +
  geom_hline(yintercept = -log10(0.05), color = "red", linetype = "dashed", size = 1.2) +
  geom_hline(yintercept = sig_bon, color = "blue", linetype = "dashed", size = 1) +
  theme(legend.position = "none") +
  scale_colour_manual(values = setNames(c('seagreen','black'), c(T,F))) +
  geom_text(x = -30, y = (-log10(0.05)-0.25), label = "p = 0.05", check_overlap = TRUE, color = "black") +
  geom_text(x = -27, y = (sig_bon-0.45), label = "p = 0.05 \n Bonferroni corrected", check_overlap = TRUE, color = "black") +
  geom_label_repel(aes(label = ifelse(-log10(p) > 5.5 | beta < -20, gene, "")), 
                   nudge_x = -0.5, nudge_y = 0.05, min.segment.length = 0) +
  labs(
    title = "PACC ~ Gene*Sex*Amyloid Burden Non-Adjusted P-Values"
  )

#plot without outlier
ggplot(subset(PACCTable_GeneSexAb, beta > -20), aes(x = beta, y = -log10(p), color = Sig)) +
  geom_point() +
  geom_hline(yintercept = -log10(0.05), color = "red", linetype = "dashed", size = 1.2) +
  geom_hline(yintercept = sig_bon, color = "blue", linetype = "dashed", size = 1) +
  theme(legend.position = "none") +
  scale_colour_manual(values = setNames(c('seagreen','black'), c(T,F))) +
  geom_text(x = -13, y = (-log10(0.05)-0.35), label = "p = 0.05", check_overlap = TRUE, color = "black") +
  geom_text(x = -12, y = (sig_bon-0.5), label = "p = 0.05 \n Bonferroni corrected", check_overlap = TRUE, color = "black") +
  geom_label_repel(aes(label = ifelse(-log10(p) > 6, gene, "")), 
                   nudge_x = -0.9, nudge_y = 0.05, min.segment.length = 0) +
  labs(
    title = "PACC ~ Gene*Sex*Amyloid Burden Non-Adjusted P-Values"
  )
```

### Summary Table {.tabset}

```{r}
datatable(subset(PACCTable_GeneSexAb, fdr_p < 0.05),
          filter = "top",
          rownames = FALSE,
          extensions = c('Buttons'),
          options = list(dom = 'Btip',
                         buttons = c('copy', 'csv'),
                         pageLength = 15,
                         columnDefs = list(list(className = 'dt-center', targets = "_all")))) %>%
  formatRound(columns = c("beta","se", "t", "p", "fdr_p"), digits = 4) %>%
  formatStyle("fdr_p", color = styleInterval(0.05, c("red", "black")))
```

### Significant Results Plots

Plots of select significant genes

::: panel-tabset
#### ENSG00000176371

```{r, fig.width = 12}
fit <- lme(PACC ~ ENSG00000176371*bl_Amyloid*SEX*pacc_time + bl_pacc_age*pacc_time + EDCCNTU*pacc_time + Cohort*pacc_time + QSVERSION + Cumulative_Dose_Scaled, 
           data = rna, random =~ pacc_time | BID, na.action = "na.omit", 
           control = list(opt = 'optim', msMaxIter = 1000, msMaxEval = 1000))


p <- plot_model(fit, type = "pred", terms = c("pacc_time", "ENSG00000176371", "bl_Amyloid", "SEX"))

p[[1]] + 
  theme(strip.background = element_rect(fill = "white")) +
  labs(
    x = "Time (Years)",
    y = "Longitudinal PACC",
    title = bquote("Longitudinal PACC by ZSCAN2, Sex, and Amyloid Status"),
    color = "ENSG00000176371\n(ZSCAN2)"
  ) +
p[[2]] + 
  theme(strip.background = element_rect(fill = "white")) +
  labs(
    x = "Time (Years)",
    y = "Longitudinal PACC",
    color = "ENSG00000176371\n(ZSCAN2)"
  )
```

#### ENSG00000163082

```{r, fig.width = 12}
fit <- lme(PACC ~ ENSG00000163082*bl_Amyloid*SEX*pacc_time + bl_pacc_age*pacc_time + EDCCNTU*pacc_time + Cohort*pacc_time + QSVERSION + Cumulative_Dose_Scaled, 
           data = rna, random =~ pacc_time | BID, na.action = "na.omit", 
           control = list(opt = 'optim', msMaxIter = 1000, msMaxEval = 1000))


p <- plot_model(fit, type = "pred", terms = c("pacc_time", "ENSG00000163082", "bl_Amyloid", "SEX"))

p[[1]] + 
  theme(strip.background = element_rect(fill = "white")) +
  labs(
    x = "Time (Years)",
    y = "Longitudinal PACC",
    title = bquote("Longitudinal PACC by SGPP2, Sex, and Amyloid Status"),
    color = "ENSG00000163082\n(SGPP2)"
  ) +
p[[2]] + 
  theme(strip.background = element_rect(fill = "white")) +
  labs(
    x = "Time (Years)",
    y = "Longitudinal PACC",
    color = "ENSG00000163082\n(SGPP2)"
  )
```
:::
